---
title: "Introduction"
---

The idea is this ...

There would be pre-written code, with explanations of what it does. Fellows would work through this notebook, running pre-written code and adding new code of their own in mini "on your own" quests.

They can Render the page as we go along to see how that works.

Practice is in a different [notebook](p1-intro.qmd).

## Libraries

Talk about them. This is fenced to show the chunk options.

```{r}
#| label: setup
#| message: false
#| echo: fenced

library(tidyverse)
library(janitor)
```


::: callout-important
There is no OYO for setting up libraries since we don't want to do it twice in the same notebook..
:::


## Importing

A lot of talk about how this works ...

```{r}
tx <- read_csv("data-raw/tx.csv") |> clean_names()

tx |> glimpse()
```


### OYO: Import

Insert a new code block here and import the weather data from a different state from the `data-raw/` folder.




> The challenge here is the page will stop rendering if they get an error, but hopefully we'll work through that since it will be an issue in any event.


## Select columns

Info about doing it. Example with Texas.

### OYO: Select

Directions to do on your own with another state.




