---
title: "Data Wrangling"
author: "Christian McDonald"
---

## Goals of this lesson

For this lesson we'll learn about data wrangling functions, many of them from [dplyr](https://dplyr.tidyverse.org/index.html). These are the functions that are much like you do in spreadsheets, like filtering, pivoting and the like. Most of these skills are necessary to prepare data before you make charts.

Functions we will learn about include: `arrange()`, `filter()`, `slice()`, `group_by()` and `summarize()`.

We'll use these to find several findings from our data, including:

- The coldest and warmest days
- The rainiest and snowiest days
- Years with most snow days
- Years with most 100+ days
- Years with most rain
- The earliest day to reach 100+ each year

To perhaps avoid some confusion we'll use just the Texas data for this lesson. I'll still be giving you your own mini-quests.

(You theoretically could use a your state, but you would need to adjust your code to import that data, use valid cities, etc.)

## Clean up our workspace

Again I have a practice notebook ready for you to fill in to keep you on track. Let's open the file and clean up our environment before we get going.

1. Open your `chjr23` project if it isn't already.
2. Open the file `practice-day1-analysis.qmd`.
3. Under the Run menu at the top of the page, choose "Restart R and Clear Output".
4. Check your Environment tab. If there is anything listed there, click on the **broom icon** to clear it out.

We do this so we don't have any leftovers from our previous notebook. Each notebook should run independently. That's also necessary to Render pages within a project.

## Add your setup chunk

I'm going to include the whole setup chunk code here again so you get the execution options.

```{r}
#| label: setup
#| message: false
#| echo: fenced

library(tidyverse)
```

We only need the tidyverse library for this notebook.

## Import our cleaned data

Now to reap the benefit our the hard work from the last lesson, let's import our cleaned data.

1. In the import section of the notebook, add a code chunk.
2. Add the `read_rds()` function below and fill out the path to your cleaned data, as indicated and run it to make sure it works.
3. Save that data into a new object called `tx_clean` and run it again. Make sure it shows up in your Environment tab.

```{r}
tx_clean <- read_rds("data-processed/tx_clean.rds")
```

## Arrange

The [arrange()](https://dplyr.tidyverse.org/reference/arrange.html) function is what we use to sort our data. We can use this function to find a couple of answers were are looking for, like what are the hottest and coldest days in our data.

The function is pretty simple, just feed it the data (usually from a pipe) and then add the column you want to sort on. By default it is in "ascending" order: low to high or alphabetically. If you want the opposite (and journalists usually do) you have to wrap the column in another function, `desc()`.

1. In the arrange section, add a code chunk for the "Coldest day" section.
2. Start with your data, the pipe into `arrange()` and add the `tmin()` column.
3. **Run the chunk** so you can see the result. Note the data is ordered by the tmin column, but it's hard to see. Let's add a `select()` function to focus on what we care about.
4. Pipe into `select()` adding the `city`, `date` and `tmin` columns.

```{r}
tx_clean |> 
  arrange(tmin) |> 
  select(city, date, tmin)
```

Glad I was not in Austin in 1949. Now to find the hottest days.

1. After the "Hottest day" header add a new code chunk.
2. Use arrange to find the highest `tmax` value. **Run** the chunk to make sure it worked.
3. Use select to clean up the fields.

```{r}
tx_clean |>
  arrange(desc(tmax)) |> 
  select(city, date, tmax)
```

Ugh, I was in Austin on both of those days in 2000 and 2011.

### OYO: Most rain

Using the same tools, find:

- The days with the most rain
- The days with the most snow

## Filter

We use the [filter()](https://dplyr.tidyverse.org/reference/filter.html) function when we want to specify rows based on a condition. This is the equivalent of clicking on the Filter tool in Excel and choosing values to keep or exclude, but we do it with code that can be fixed and repeated.

We'll use filter as one function of many for some of our answers, like which years had the most 100+ degree days. We need to work on the filtering skill here first, as there are nuances.

The function works like this:

``` r
# this is psuedo code. don't add it
data |> 
  filter(variable comparison value)

# example
tx_clean |> 
  filter(city == "Austin")
```

The `filter()` function typically works in this order:

-   What is the variable (or column) you are searching in.
-   What is the comparison you want to do. Equal to? Greater than?
-   What is the observation (or value in the data) you are looking for?

Note the two equals signs `==` in our Austin example above. It is important to use two of them when you are asking if a value is "true" or "equal to", as a single `=` typically means you are assigning a value to something.

### Comparisons: Logical tests

There are a number of these logical tests for the comparison:

| Operator          | Definition               |
|:------------------|:-------------------------|
| x **\<** y        | Less than                |
| x **\>** y        | Greater than             |
| x **==** y        | Equal to                 |
| x **\<=** y       | Less than or equal to    |
| x **\>=** y       | Greater than or equal to |
| x **!=** y        | Not equal to             |
| x **%in%** c(y,z) | In a group               |
| **is.na(**x**)**  | Is NA (missing values)   |
| **!is.na(**x**)** | Is not NA                |

Where you apply a filter matters. If we want to consider only certain data before other operations, then we need to do the filter first. In other cases we may filter after all our calculations to show just rows of interest.

### Single condition

Let's find days that are 100+.

1. In the `## Filter` section in the part about 100+ days, add a code chunk.
2. Start with the data, then pipe into `filter()`.
3. For the condition, look in the `tmax` column using `>=` to find values "greater or equal to" `100`.
4. **Run the code** to make sure it works.
5. Use `select()` to focuse on the variables of interest.

```{r}
tx_clean |> 
  filter(tmax >= 100) |> 
  select(city, date, tmax)
```

### Multiple "and" conditions

OK, that's fine, but our list is not long enough to see the days in Dallas. Let's do this again, but add a second condition to test. When you use a comma `,` or ampersand `&` between conditions, both conditions must be true to keep the rows.

1. In the Dallas 100+ section, start a new code chunk.
2. Do the same code as above and **run** it to make sure it still works.
3. Use a comma after the first condition to add a second one: `city == "Dallas"`.


```{r}
tx_clean |> 
  filter(tmax >= 100, city == "Dallas") |> 
  select(city, date, tmax)
```

### Multiple "or" conditions

But what if we have an "either or" case where one of two conditions could be true. This would be the case if we wanted to find days where it either a) snowed that day, or b) there was snow left on the ground from a previous day. This is a true snow day, right?

You can use the `|` operator for an "or" condition. That character is the shift of the `\` key just above your return/enter key.

1. In the section about snow days, add a chunk.
2. Add the code, but run after adding the first condition, before you add the second condition, so you can compare them when you are done.
3. Add the second condition and run it again.
4. Add the select to focus on our columns of interest.


```{r}
tx_clean |> 
  filter(snow > 0 | snwd > 0) |> 
  select(city, date, snow, snwd)
```


### But I need "and" and "or"

You can mix "and" and "or" conditions, but note the order of them might matter depending on what you are doing.

### OYO: Real snow days in Dallas

In your notebook, start with the snow days we had above, but add a new condition to the filter that filters `city` to just "Dallas".

### Filtering text

We don't really have a need for this in our data, but we can use filter to find parts of words as well. There are many ways, but the one I use the most is `str_detect()`.

```{r}
tx_clean |> 
  filter(str_detect(city, "st")) |> # <1>
  distinct(city) # <2>
```

1. Inside the filter, we start with `str_detect()`. The first argument it needs is which column in our data to look at, so we fed it `city`. The second argument (after a comma) is the string of text we are looking for in the column, which is `st` in our case.
2. I used `distinct(city)` here so we could more easily see our results.

The code above found both "Hou**st**on" and "Au**st**in" because they have "st" in them. It didn't capture "Dallas".

## Slice

Another way to pick out specific rows of data based on a condition is to use [slice](https://dplyr.tidyverse.org/reference/slice.html) variables like `slice_max()`, `slice_min()`.

Let's use `slice_min()` to find the coldest day in our data.

1. In the slice section of the notebook about coldest day, add the following:

```{r}
tx_clean |> 
  slice_min(tmin) |> 
  select(city, date, tmin)
```


We get one result, the coldest day in the data. But what if we want the coldest day for each city? We will introduce `group_by()` to solve that.

## Group by

The `group_by()` function goes behind the scenes to organize your data into groups, and any function that follows it gets executed **within** those groups.

The columns you feed into `group_by()` determine the groups. If we do `group_by(city)` then  all the "Austin" rows are grouped together, then all the "Dallas" rows, then all the "Houston" rows.

I sometimes think of these groups as piles of data, separate from each other. We would have three piles of data, one for each city. Functions that follow happen independently on each pile.

## Group and slice

If we add our `group_by(city)` before slice, then it works **within** each group. Like this:

```{r}
tx_clean |> 
  group_by(city) |> 
  slice_min(tmin) |> 
  select(city, date, tmin)
```

Look at the difference in this result. Now we get a result for each city, because the rows we "grouped" the data **before** performing the slice. Since there are three cities, we get three results.

## Multiple groupings

We can also group by multiple columns. What that does is create a group (or pile!) of data for each matching combination of values.

So, if we `group_by(city, yr)` then we will get a pile for each year of Austin (85 piles because there are 85 years of data for Austin), then a pile for each year in Houston, etc.

If we were to find the hottest day in each of those piles, it would look like this:

1. Create a new chunk and add this to your notebook using copy-to-clipboard.
2. Try it with and without the `distinct()` at the end and think about why you got those results.

```{r}
tx_clean |> 
  group_by(yr, city) |> 
  slice_max(tmax) |> 
  select(city, tmax) |> 
  distinct()
```

I added distinct the `distinct()` there to remove some ties where there were multiple days in a year with that same high temperature.

## Summarize

While slice is nice, we really went through this exercise to understand `group_by()` so we can use it with [summarize()](https://dplyr.tidyverse.org/reference/summarise.html), which allows us to summarize data much like a pivot table in Excel.

::: callout-tip
`summarise()` and `summarize()` are the same function. The creator of tidyverse is from New Zealand so he has both spellings. I tend to use them both by whim, though the "s" version comes first in type-assist.
:::

If there are no group_by variables, the output of `summarize()` will be a single row summarizing all observations in the input. If we have groups, it will contain one column for each grouping variable and one column for each summary statistic we specify. Much easier to understand if we do it.

Let's first do one without groups.

1. In the Summarize section, add the following chunk and code.

```{r}
tx_clean |> 
  summarize(
    e_date = min(date),
    l_date = max(date),
    cnt = n()
  )
```

We have no groups here, so we just get the stats we specified ... the earliest date in our data, the latest date in our data and the number of rows.

**It is important** to name our summarized columns like the `e_date =` part above. If we don't, the function becomes a new column name it is hard to work with later. We can name the summarized column anything we want, but choose something that _describes_ the column.

### Add city as a group

1. Just use the copy-to-clipboard tool to add this to your notebook and run it to see it.

```{r}
tx_clean |> 
  group_by(city) |> # <1>
  summarise(
    e_date = min(date),
    l_date = max(date),
    cnt = n()
  )
```

1. This is where we add the group.

Now that we've added a group, we get a result for each unique value within that column. We have three cities, so we get three results.

### Add both city and yr as a group

If we group by multiple things, then again we'll get results for each unique combination between the two variables.

1. Just use the copy-to-clipboard tool to add this to your notebook and run it to see it.

```{r}
tx_clean |> 
  group_by(city, yr) |> # <1>
  summarise(
    e_date = min(date),
    l_date = max(date),
    cnt = n()
  )
```

1. This is where we add the second group to get both city and yr.

Now we get a result for each year within each city.

You'll also see this message about "`summarise()` has grouped output by 'city'." That is not an error, it's just telling you that your data will remain grouped by city if you add more functions. It's a rabbit hole we won't jump down today, but know it is not a problem.

### Common summarize stats

There are a number of statistics we can find about our data within summarize.

- `n()` counts the number of rows
- `n_distinct()` counts the number of distinct values in the column
- `min()` gives the smallest value
- `max()` gives the largest value

Some math operators might need the argument `na.rm = TRUE` to ignore NA (empty) values.

- `sum()` adds values together.
- `mean()` gives the mean (or average) value.
- `median()` gives the median or middle value

There are other useful ones in the [summarise documentation](https://dplyr.tidyverse.org/reference/summarise.html).

## Group and summarize: Count

A very typical workflow to answer a data-driven question is to count records. Often the logic is to:

- Do we need to consider all the data or just some of it?
- Group the records by variables of interest, like categories or dates.
- Count the number of records in each group.

We'll use these general steps to answer this question: **What years have had the most 100+ degree days.** We'll start with Austin with careful consideration, then generally show how do it for all the cities.

Here are the steps I used to do this, along with my thought process. In some cases I'm adding code to check something and then deleting it before moving in, but the code below just shows the end result.

1. I started with a new code chunk with the `tx_clean` data.
2. I then filtered it to Austin and ran it to check it. I saved that result into a new object, `atx`.
3. I added a line that used `summary()` to check the dates of the `atx` data. It looks like the data starts in June 1938 and there could have been days before that that were 100, so I amended my filter to cut out 1938. The latest date is Sept. 30. 2023. Since this is our current year and there are typically few 100+ days in October, I'll keep this year, but I'll note it if I use this in a chart.
4. I glimpse it again just for convenience to see the column names.

Here is the code:

```{r}
# get Austin data
atx <- tx_clean |> filter(city == "Austin", yr > 1938)

# check dates
atx$date |> summary()

# peek
atx |> glimpse()
```

With my prepped data I can now do my calculations:

1. I start with the `atx` data and filter to get days where `tmax` was 100+. I **Run the chunk** to check it.
2. Then I grouped the records by `yr`.
3. Then I summarize to get the count using `n()`, but with a descriptive column name, `hot_days`.
4. Then I arranged the results to show the most `hot_days` at the top.
5. Then I added a filter to cut off the list at a logical place, years with 50 days or more.

```{r}
atx |> 
  filter(tmax >= 100) |> 
  group_by(yr) |> 
  summarize(hot_days = n()) |> 
  arrange(desc(hot_days)) |> 
  filter(hot_days >= 50)
```

Here is what that looks like if I try to do it with all the cities. The difference here is I'm not taking as much care with the dates so there might be partial years that are undercounted, but they should at least be lower.

```{r}
tx_clean |> 
  filter(tmax >= 100) |> 
  group_by(yr, city) |> 
  summarize(hot_days = n()) |> 
  arrange(desc(hot_days)) |> 
  filter(hot_days >= 30)
```

Apparently the Houston Hobby Airport just doesn't have that many 100+ degree days compared to Austin and Dallas. That humidity, though ...

### OYO: Most snow days by city each year

If there is time, you could use a similar method to count the number of days with with `snow`.


## Group and Summarize: Math

The next question we want to answer: **In each city, which years had the most rain and which had the least?** Let's walk through our logic questions again:

_Do we need to consider all the data?_

In this case we definitely want only whole years. When we tried summarize above we calculated the first date of all the data, and looking at that we'll have to start with 1940. We'll also lop off 2023. 

_Do we need to consider our data in any groups?_

We need to add together values in each city and year, so Austin for 1940, then '41, etc. We'll group our data by `city` and `yr`.

_What do we need to calculate?_

We need to add together the inches of rain, so we can `sum()` the `rain` column.

Work through this in your practice notebook step by step:

1. In the "Group and Summarize: Math" section, add a code chunk.
2. Start with our data, then filter it to start after 1939 and before 2023. **Run** the chunk to make sure it works. You might even pipe into a `slice_max(date)` to test if it worked (but then remove it after you have checked.)
3. Group your data by `city` and `yr`.
4. Summarize your data using `sum()` on the `rain` column. You probably have to add an argument `na.rm = TRUE` to make this work because some days have blank values for rain and `sum()` doesn't know what to do with those blank values. Run it!
5. Arrange the values first by city, then by the summed rain in descending order. Run it!
6. Save your result into an object and then print it back out so you can see it. Run it!

```{r}
tx_yr_rain <- tx_clean |> 
  filter(yr > 1939, yr < 2023) |>
  group_by(city, yr) |> 
  summarise(tot_rain = sum(rain, na.rm = TRUE)) |> 
  arrange(city, desc(tot_rain))

tx_yr_rain
```

At this point we have all the values, through they are hard to read. Some paths we could take to get more clarity:

- We could take our new object and build new blocks to filter by city and arrange by most or least rain. Six new code chunks. Very clear.
- We could maybe use group and slice_max to find highest values within each city, and then again with slice_min. There is an [option](https://dplyr.tidyverse.org/reference/slice.html) to set the number of returns.
- We could plot this on a chart (a lesson for another day!). 

Here is the group and slice method where I use the `n = ` argument to set how many records I want in the slice.

```{r}
tx_yr_rain |> 
  group_by(city) |> 
  slice_max(tot_rain, n = 3)
```

Here is the least rain:

```{r}
tx_yr_rain |> 
  group_by(city) |> 
  slice_min(tot_rain, n = 3)
```

### OYO: Years with most snow

Try to do the same to find the years with the most snow in each city.

## Working through logic

Here is a question that takes a couple of steps to accomplish: **What is the average monthly rainfall for each city?** i.e., How much rain should we expect each month, based on history?

_Do we need to consider all our data?_ We don't want partial years or months or our math won't work out right, so again, we want just full years, so 1940 through 2022 for the Texas data.

_Do we need to consider data in groups?_ This is tricky. We can't just group by month and get the average because then we would be averaging the rain _each_ day within a month. We have to total the rain within a month for each year, then we can get the average.

_What calculations do we need?_ Kinda answered that in that we need two calculations: One to total the rain within each month/year, and another to get the averages across those months.

We could do the code all in one chunk, but you wouldn't see the result of the first group and sum, so we'll do it in two.

### First: Sum rain within a city, month, year

1. Use the copy-to-clipboard option to add this to your notebook in the "Working through logic" section. Annotations about the code follow.

```{r}
#| message: false

tx_mn_yr_rain <- tx_clean |> 
  filter(yr >= 1940, yr <= 2022) |> # <1>
  group_by(city, mn, yr) |>  # <2>
  summarize(mn_yr_rain = sum(rain, na.rm = TRUE)) # <3>

tx_mn_yr_rain  
```

1. We filter to get our full years.
2. We group by three things so we can add the rain in each city for each month of each year.
3. We then sum the rain with a nice name. We use `na.rm = TRUE` because there were cases where the rain value was blank instead of 0.0. Perhaps the station was down? Would need to do some reporting to figure that out.

### Second: Average rain by city, month

Now we can group by city and month to average together all the "Jan" in "Austin", then all the "Feb" in "Austin", etc.

1. Use the copy-to-clipboard option to add this to your notebook. Annotations follow.

```{r}
city_avg_rain <- tx_mn_yr_rain |> 
  group_by(city, mn) |> # <1>
  summarise(avg_mn_rain = mean(mn_yr_rain)) # <2>

city_avg_rain
```

1. We take the result from above and group now by each city and each month, so we can work with all the "Austin in Jan" to get our average.
2. Now we can get the average `mean()` of those months.

### Plot the results

And to tease you to what is in store for you tomorrow, let's plot those results. We won't get into each line of code here, I just want to show you it can be done.

```{r}
city_avg_rain |> 
  ggplot(aes(x = mn, y = avg_mn_rain, group = city)) +
  geom_line(aes(color = city)) +
  ylim(0,6) +
  labs(
    title = "Average monthly rainfall, 1940-2022",
    x = "", y = "Average monthly rain",
    color = "City"
  )
```


## Challenge: Earliest 100+ day each city

Find the earliest date in each city where it reached at least 100 degrees. Here is a hint: Slice is nice and we have the `yd` field that is the number of days into a year.

## What do you want to learn about the weather?

Dream up a question and answer it. Maybe import your own state's data and find something there. The earliest snowfall. The snowiest month on average. The first freeze. The last freeze. Or more challenging: the average day of the last freeze (i.e., when can I plant my garden!).
