---
title: "Practice Day 2"
---

## Goals

To learn about: arrange, filter, slice, group_by, summarize

To find several values from our data:

- The coldest and warmest days
- The rainiest and snowiest days
- Years with most snow days
- Years with most 100+ days
- Years with most rain
- Earliest day to reach 100+ each year

With this lesson we'll just use Texas data. (You theoretically could use a different state, but would need to adjust your code to import the right data, use valid cities, etc.)

## Setup

```{r}
#| label: setup
#| message: false

library(tidyverse)
```

## Import

Import your cleaned data using `read_rds()` and save it into an object:

```{r}

```

## Arrange

Find the coldest day, warmest day, most snow, most rain.

### Coldest day

Find the coldest day:

```{r}

```

### Hotest day

Find the hottest day:

```{r}

```

### OYO: Most rain

Find the days with the most rain:

```{r}

```

### OYO: Most snow

Find the days with the most snow:

```{r}

```

## Filter

Find days that are 100+:

```{r}

```

Filter for days in Dallas that are 100+:

```{r}

```

Find days where it snowed, or there is snow still on the ground:

```{r}

```

### OYO: Snow days in Dallas

Find days where it snowed or there is snow on the ground, but only in Dallas.

```{r}

```

## Slice

Use slice_min to find the coldest day in our data:

```{r}

```

## Group and slice

Add `group_by` to find the coldest day in each city.

```{r}

```

### OYO: Hottest day in each city

Use `group_by` and `slice_max` to find the hottest days in each city. Note there might be some ties.

```{r}

```

### Multiple groups

Hottest day each year in each city:

```{r}

```

## Summarize

Summarize to find our first date, last date and number of rows:

```{r}

```

## Group and summarize

Group the data by city and find the first date, last date and number of rows:

```{r}

```

Add city and yr as a group:

```{r}

```


## Group and summarize: Count

Find the number of days in Austin that were 100+. Start first with prepping the data to get the date ranges we need:

```{r}

```

Then do the calculation to get hot days:

```{r}

```

Find the years with the most 100+ degree days in each city.

```{r}

```

### OYO: Most snow days by city each year

Count only the days that where it snowed.

```{r}

```

## Group and Summarize: Math

Years with most rain in each city.

```{r}

```

The most rain in each city, sliced:

```{r}

```

The least rain in each city, sliced:

```{r}

```

### OYO: Years with most snow

Find the years with the most total snow in each city

```{r}

```

Most snow, sliced:

```{r}

```


## Working through logic

Getting average monthly rain for each city.

First get the total rain for each month/year:

```{r}
#| message: false

```

Then calculate the average for the months in each city:

```{r}
#| message: false

```

And as a tease, we plot it:

```{r}

```


## Challenge: Earliest 100+ day each city

For each city, find the earliest day of a year in which it reached 100 degrees.

```{r}

```
