---
title: "Solutions Day 1"
---

This is the notebook fellows will work through on the first day of the Center for Health Journalism Hands-On R course.

You will be using daily weather summaries that have been downloaded  from [Climate Data Online](https://www.ncei.noaa.gov/cdo-web/datasets). The explanations use Texas, but there are files for Arkansas, California, New York and North Carolina for practice.

## Goals

Our goals are to:

- Import our data.
- Check all the column data types.
- Add some new columns based on the date.
- Recode some values in our data.
- Remove some unnecessary variables/columns.
- Export our cleaned data.

## Setup

Add the entire code block for libraries.

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(janitor)
```

## Import

Follow the directions in the lesson to import the Texas data, starting with adding a new code block:

```{r}
tx_raw <- read_csv("data-raw/tx.csv") |> clean_names()

tx_raw
```

### OYO: Import a different state

Go through all the steps above, but with different a different state.



## Peeking at data

Use head, tail, glimpse and summary to look at the Texas data.

Look at the top of your data:

```{r}
tx_raw |> head()
```

Look at 8 lines of the bottom of your data:

```{r}
tx_raw |> tail(8)
```


Use glimpse to see all your columns:

```{r}
tx_raw |> glimpse()
```

Use summary to learn about all your variables:

```{r}
tx_raw |> summary()
```

```{r}
tx_raw$date |> summary()
```

### OYO: Peek at your state's data



## Create or change data

Create year, month values based on the date.

```{r}
tx_dates <- tx_raw |> 
  mutate(
    yr = year(date),
    mn = month(date, label = TRUE),
    yd = yday(date)
  )

tx_dates |> glimpse()
```

### OYO: Make date parts

Make the same date parts, but with your own state data:


## Recoding values

Use distinct so you can see the station names:

```{r}
tx_dates |> distinct(name)
```

### Use mutate to recode

Use recode to create a new column of short city names:

```{r}
tx_names <- tx_dates |> 
  mutate(
    city = recode(
      name,
      "HOUSTON WILLIAM P HOBBY AIRPORT, TX US" = "Houston",
      "AUSTIN CAMP MABRY, TX US" = "Austin",
      "DALLAS FAA AIRPORT, TX US" = "Dallas"
    )
  )

tx_names |> glimpse()
```

Now check your results using distinct on `name` and `city`.

```{r}
tx_names |> distinct(name, city)
```


### OYO: Recode your cities

Make similar short names, but for your state.


## Select

Create a new version of your data with only the columns you need, in the order you want them.

```{r}
tx_tight <- tx_names |> 
  select(
    city,
    date,
    rain = prcp,
    snow,
    snwd,
    tmax,
    tmin,
    yr,
    mn,
    yd
  )

tx_tight |> glimpse()
```

### OYO: Select your cols

Go through the same process as above, but with your own state data.



## Export

Write the file out as "rds" to the `data-processed` folder.

```{r}
tx_tight |> write_rds("data-processed/tx_clean.rds")
```

### OYO: Export your state

Write your data to the `data-processed` folder. Make sure you use a name for your state.


## Checking your notebooks

Clear out your notebook and rerun all the code. Render the HTML page.

