---
title: "Weather testing"
---


## California - R
- Los Angeles - Downtown USC
- Sacramento - Metro airport
- San Francisco - Downtown

## North Carolina - R
- Charlotte Douglas Airport
- Raleigh Durham International Airport
- Wilmington International Airport
Texas
- Houston William P Hobby
- Dallas FAA
- Austin - Mabry

## Arkansas
- Little Rock
- Fort Smith
- Hot Springs

## New York -R
- NY City Central Park
- Albany International Airport
- Buffalo

## Stations

I really only need TX, CA & NC

- AR: GHCND:USW00013963, GHCND:USC00033466, GHCND:USW00013964
- CA: GHCND:USW00093134, GHCND:USW00093225, GHCND:USW00023272
- NC: GHCND:USW00013881, GHCND:USW00013748, GHCND:USW00013722
- TX: GHCND:USW00013960, GHCND:USW00012918, GHCND:USW00013958
- NY: GHCND:USW00094728, GHCND:USW00014735, GHCND:USW00014733

### CA notes

DAPR = Number of days included in the multiday precipitation total (MDPR)
MDPR = Multiday precipitation total (mm or inches as per user preference; use with DAPR and DWPR, if available)

Testing the weather data

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(janitor)
```

## Import

```{r}
ca_raw <- read_csv("../data-raw/ca.csv") |> clean_names()
tx_raw <- read_csv("../data-raw/tx.csv") |> clean_names()
nc_raw <- read_csv("../data-raw/nc.csv") |> clean_names()
```



```{r}
ca <- ca_raw |> select(
  name, date, prcp, snow, snwd, tmax, tmin
)

tx <- tx_raw |> select(
  name, date, prcp, snow, snwd, tmax, tmin
)

nc <- nc_raw |> select(
  name, date, prcp, snow, snwd, tmax, tmin
)


```

