---
title: "Weather testing"
---

## About this notebook

Here we look at all the weather data gathered here, checking the dates and such.

The data are [Global Historical Climate Network Daily Summaries](https://www.ncei.noaa.gov/cdo-web/search?datasetid=GHCND).

## Stations

These are the station identifiers:

- AR: USW00013963, USC00033466, USW00013964
- CA: USW00093134, USW00023271, USW00023272
- NC: USW00013881, USW00013748, USW00013722
- TX: USW00013960, USW00012918, USW00013958
- NY: USW00094728, USW00014735, USW00014733

I really only need TX, CA & NC, I think.

### Extra column notes

- DAPR = Number of days included in the multiday precipitation total (MDPR)
- MDPR = Multiday precipitation total (mm or inches as per user preference; use with DAPR and DWPR, if available)
- DASF = Number of days included in the multiday snowfall total (MDSF)
- MDSF = Multiday snowfall total (mm or inches as per user preference)
- TOBS = Temperature at the time of observation (Fahrenheit or Celsius as per user preference)
- TAVG = Average of hourly values

Testing the weather data

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(janitor)
```

## Import

```{r}
ca_raw <- read_csv("../data-raw/ca.csv") |> clean_names()
tx_raw <- read_csv("../data-raw/tx.csv") |> clean_names()
nc_raw <- read_csv("../data-raw/nc.csv") |> clean_names()
ny_raw <- read_csv("../data-raw/ny.csv") |> clean_names()
ar_raw <- read_csv("../data-raw/ar.csv") |> clean_names()
```

## Bind data

```{r}
wx_raw <- ar_raw |> 
  bind_rows(ca_raw, nc_raw, ny_raw, tx_raw)

wx_raw |> glimpse()
```

## Add state

Get the state name from the station name.

```{r}
wx_states <- wx_raw |> 
  mutate(
    state = str_sub(name, -5, -3)
  )

wx_states |> glimpse()
```

## Recode for city

Get the names:

```{r}
wx_states |> 
  distinct(name)
```

Recode for city names

```{r}
wx_cities <- wx_states |> 
  mutate(city = recode(
    name,
    "FORT SMITH REGIONAL AIRPORT, AR US" = "Fort Smith",
    "LITTLE ROCK AIRPORT ADAMS FIELD, AR US" = "Little Rock",
    "HOT SPRINGS 1 NNE, AR US" = "Hot Springs",
    "SAN FRANCISCO DOWNTOWN, CA US" = "San Francisco",
    "LOS ANGELES DOWNTOWN USC, CA US" = "Los Angeles",
    "SACRAMENTO 5 ESE, CA US" = "Sacramento",
    "CHARLOTTE DOUGLAS AIRPORT, NC US" = "Charlotte",
    "WILMINGTON INTERNATIONAL AIRPORT, NC US" = "Wilmington",
    "RALEIGH DURHAM INTERNATIONAL AIRPORT, NC US" = "Raleigh",
    "NY CITY CENTRAL PARK, NY US" = "New York City",
    "BUFFALO NIAGARA INTERNATIONAL, NY US" = "Buffalo",
    "ALBANY INTERNATIONAL AIRPORT, NY US" = "Albany",
    "HOUSTON WILLIAM P HOBBY AIRPORT, TX US" = "Houston",
    "AUSTIN CAMP MABRY, TX US" = "Austin",
    "DALLAS FAA AIRPORT, TX US" = "Dallas",
  ))

wx_cities |> distinct(name, city)
```



## Select cols

```{r}
wx_tight <- wx_cities |> 
  select(
    name,
    city,
    state,
    date,
    prcp,
    snow,
    snwd,
    tmax,
    tmin
  )

wx_tight |> glimpse()
```

## Check the dates

```{r}
wx_tight |> 
  group_by(city, state) |> 
  summarise(
    f_date = min(date),
    l_date = max(date),
    cnt = n()
  ) |> 
  arrange(state, f_date)
```

